---
title: "Stats-506 Problem Set #3"
format: pdf
author: "Zhekai Liu"
subtitle: "[GitHub Repository: https://github.com/zkl2002/Stats-506/](https://github.com/zkl2002/Stats-506/)"
editor: visual
---

```{r, message=FALSE, warning=FALSE}
library(haven)
library(dplyr) 
library(knitr)
library(DBI)
library(RSQLite)
library(microbenchmark)
library(pscl)
```

# Problem 1

## (a).

```{r}
# read files
aux  <- read_xpt("AUX_I.XPT")
demo <- read_xpt("DEMO_I.XPT") 
# merge data set
df <- inner_join(aux, demo, by = "SEQN")
dim(df)
```

The merged date frame has 4582 rows and 119 columns.

## (b).

```{r}
# switch missing code to NA
df$DMDCITZN[df$DMDCITZN %in% c(7, 9)] <- NA_integer_
df$INDHHIN2[df$INDHHIN2 %in% c(77, 99)] <- NA_integer_

# gender as factor
df$gender <- factor(df$RIAGENDR, levels = c(1, 2),
                         labels = c("Male", "Female"))
# citizenship as factor
df$citizen <- factor(df$DMDCITZN, levels = c(1, 2),
                         labels = c("Citizen", "Not citizen"))
# children under 5 years old
df$children_under5 <- factor(df$DMDHHSZA, 
                             levels = c(0, 1, 2, 3), 
                             labels = c("0", "1", "2", "3 or more"),
                             ordered = TRUE)

# household income
codes  <- c(1:10, 12, 13, 14, 15)
labels <- c("$0 to $4,999", 
            "$5,000 to $9,999",
            "$10,000 to $14,999",
            "$15,000 to $19,999",
            "$20,000 to $24,999",
            "$25,000 to $34,999",
            "$35,000 to $44,999",
            "$45,000 to $54,999",
            "$55,000 to $64,999",
            "$65,000 to $74,999",
            "$20,000 and Over",
            "Under $20,000",
            "$75,000 to $99,999",
            "$100,000 and Over")

df$household_income <- factor(df$INDHHIN2, levels = codes,
                              labels = labels, ordered = TRUE)

df_clean <- df[c("SEQN", "gender", "citizen", 
                 "children_under5", "household_income")]
```

`df_clean` is the clean up data frame and all categorical values converted to factor with informative values.

## (c).

```{r}
# right ear
xR <- df$AUXTWIDR
xR[xR %in% c(555, 777, 888)] <- NA_integer_
xR[xR < 0 | xR > 500] <- NA_integer_
df_clean$tw_right <- as.integer(xR)

# left ear
xL <- df$AUXTWIDL
xL[xL %in% c(555, 777, 888)] <- NA_integer_
xL[xL < 0 | xL > 500] <- NA_integer_
df_clean$tw_left <- as.integer(xL)

# children number as continuous
df_clean$children <- df$DMDHHSZA
df_clean$children <- as.integer(df_clean$children)

# income as continuous
inc <- df$INDHHIN2
inc[inc %in% c(77, 99)] <- NA_integer_ 

codes_keep <- c(1:10, 12:15)
mids <- c( 2500, 7500, 12500, 17500, 22500, 30000,
          40000, 50000, 60000, 70000, 20000, 10000,
          87500, 100000)

df_clean$income <- NA_real_
idx <- match(inc, codes_keep)
df_clean$income[!is.na(idx)] <- mids[idx[!is.na(idx)]]
```

We treat `children` as a continuous count by coding “3 or more” as 3, and we convert `household income` to a continuous measure using category midpoints (codes 77/99 set to missing). If some household income is over 20,000 or over 100,000, we just set as 20,000 and 100,000.

```{r}
# models
m1R <- glm(tw_right ~ gender, data = df_clean, 
           family = poisson, na.action = na.omit)

m2R <- glm(tw_right ~ gender + citizen + children + income, 
           data = df_clean, family = poisson, na.action = na.omit)

m1L <- glm(tw_left  ~ gender, data = df_clean, 
           family = poisson, na.action = na.omit)

m2L <- glm(tw_left  ~ gender + citizen + children + income, 
           data = df_clean, family = poisson, na.action = na.omit)
```

```{r}
b1R <- exp(coef(m1R))
b2R <- exp(coef(m2R))
b1L <- exp(coef(m1L))
b2L <- exp(coef(m2L))

ref_len <- length(b2R)
coeftable <- cbind(
  c(b1R, rep(NA, ref_len - length(b1R))),
  c(b2R, rep(NA, ref_len - length(b2R))),
  c(b1L, rep(NA, ref_len - length(b1L))),
  c(b2L, rep(NA, ref_len - length(b2L)))
)
rownames(coeftable) <- names(b2R)
colnames(coeftable) <- c("1R","2R","1L","2L")

coeftable <- round(coeftable, 3)
mode(coeftable) <- "character"; coeftable[is.na(coeftable)] <- ""
kable(coeftable, align = "cccc", caption = "IRR (exp(beta)) including baseline")

```

```{r}
N <- c(nobs(m1R), nobs(m2R), nobs(m1L), nobs(m2L))

pseudo <- c(pR2(m1R)["McFadden"],
            pR2(m2R)["McFadden"],
            pR2(m1L)["McFadden"],
            pR2(m2L)["McFadden"])

AICv <- c(AIC(m1R), AIC(m2R), AIC(m1L), AIC(m2L))

stats_tab <- rbind(
  "Sample size" = sprintf("%d",   N),
  "Pseudo R^2"  = sprintf("%.5f", pseudo),
  "AIC"         = sprintf("%.2f", AICv)
)

colnames(stats_tab) <- c("1R","2R","1L","2L")

knitr::kable(stats_tab, align = "lrrrr",
             caption = "Models sample size, pseudo-R², and AIC")
```

## (d).

```{r}
summary(m2L)
```

The result is statistically significant, and estimate coefficient for females vs. males is **greater than 0**. This provides evidence that, holding citizenship, number of young children, and household income constant, females have a **higher expected tympanometric width in the left ear** than males.

```{r}
keep <- complete.cases(df_clean[, c("tw_left","gender",
                                    "citizen","children","income")])
dd   <- df_clean[keep, ]

new_m <- data.frame(
  gender   = factor("Male",   levels=levels(dd$gender)),
  citizen  = factor("Citizen",levels=levels(dd$citizen)),
  children = mean(dd$children),
  income   = mean(dd$income, na.rm=TRUE)
)
new_f <- new_m
new_f$gender <- factor("Female", levels=levels(dd$gender))

mu_m <- exp(predict(m2L, new_m, type = "link"))
mu_f <- exp(predict(m2L, new_f, type = "link"))
ratio <- mu_f / mu_m

p2 <- summary(m2L)$coef["genderFemale","Pr(>|z|)"]

data.frame(
  Setting = "Citizen; children=mean; income=mean",
  Pred_Male   = round(mu_m, 2),
  Pred_Female = round(mu_f, 2),
  Ratio_F_over_M = round(ratio, 3),
  p_value = signif(p2, 3)
)

```

By controlling for citizenship, number of children, and household income, females have a higher expected tympanometric width than males. The estimated **incidence rate ratio** for females vs. males is **1.015** (≈ **1.5%** higher), with $\mathbf {\text{p} = 1.95\times 10^{-5}}$ , indicating a statistically significant difference.

## **Problem 2 - Sakila**

## (a).

```{r}
sakila <- dbConnect(RSQLite::SQLite(), "sakila_master.db")
```

**R code method:**

```{r}
# Output of problem a with R code
R_a <- function() {
  cust <- dbGetQuery(sakila, "SELECT customer_id,
                      store_id, active FROM customer;")
  tot <- aggregate(customer_id ~ store_id, cust, length)
  names(tot)[2] <- "n_customers"

  act <- aggregate(as.numeric(active) * 100 ~ store_id, cust, mean)
  names(act)[2] <- "active_pct"

  out <- merge(tot, act, by = "store_id", sort = TRUE)
  out
}
R_a()
```

**SQL method:**

```{r}
SQL_a <- function() {
  dbGetQuery(sakila, "
    SELECT store_id,
           COUNT(customer_id) AS n_customers,
           ROUND(AVG(active) * 100.0, 4) AS active_pct
    FROM customer
    GROUP BY store_id
    ORDER BY store_id;")
}
SQL_a()
```

```{r}
bench_a <- microbenchmark(
  R   = R_a(),
  SQL = SQL_a()
)
print(bench_a)
```

Both methods give the same result, and SQL is faster.

## (b).

**R method:**

```{r}
# Output of problem b with R code
R_b <- function(){
  staff <- dbGetQuery(sakila, "SELECT staff_id, first_name,
                        last_name, address_id FROM staff;")
  address <- dbGetQuery(sakila, "SELECT address_id, city_id FROM address;")
  city <- dbGetQuery(sakila, "SELECT city_id, country_id FROM city;")
  country <- dbGetQuery(sakila, "SELECT country_id, country FROM country;")
  staff_addr <- merge(staff, address, by = "address_id", all.x = TRUE)
  staff_city <- merge(staff_addr, city, by = "city_id", all.x = TRUE)
  staff_country <- merge(staff_city, country, by = "country_id", all.x = TRUE)
  res <- staff_country[, c("staff_id", "first_name", "last_name", "country")]
  res <- res[order(res$staff_id), ]
  rownames(res) <- NULL
  kable(res, align = "cccc")}
R_b()
```

**SQL method:**

```{r}
SQL_b <- function(){
    res_sql <- dbGetQuery(sakila, "
    SELECT s.staff_id,
           s.first_name,
           s.last_name,
           co.country AS country
    FROM staff AS s
    JOIN address AS a ON a.address_id = s.address_id
    JOIN city    AS c ON c.city_id    = a.city_id
    JOIN country AS co ON co.country_id = c.country_id
    ORDER BY s.staff_id;
  ")
  kable(res_sql, align = "cccc")
}
SQL_b()
```

```{r}
bench_b <- microbenchmark(
  R = R_b(),
  SQL = SQL_b()
)
print(bench_b)
```

Both methods give the same result, and SQL is faster.

## (c).

```{r}
# Output of problem c with R code
R_c <- function(){
  pay <- dbGetQuery(sakila, "SELECT rental_id, amount FROM payment;")
  rent <- dbGetQuery(sakila, "SELECT rental_id, inventory_id FROM rental;")
  inv <- dbGetQuery(sakila, "SELECT inventory_id, film_id FROM inventory;")
  film <- dbGetQuery(sakila, "SELECT film_id, title FROM film;")
  sum_pay <- aggregate(amount ~ rental_id, pay, sum)
  pay_max <- max(sum_pay$amount)
  rental_max <- sum_pay$rental_id[sum_pay$amount == pay_max]
  inv_max <- unique(rent$inventory_id[rent$rental_id %in% rental_max])
  film_max <- unique(inv$film_id[inv$inventory_id %in% inv_max])
  film_names <- film$title[film$film_id %in% film_max]
  film_names
}
R_c()
```

**SQL method:**

```{r}
SQL_c <- function(){
  dbGetQuery(sakila, "
  SELECT DISTINCT f.title
  FROM (
    SELECT rental_id, SUM(amount) AS total_amount
    FROM payment
    GROUP BY rental_id
  ) AS t
  JOIN rental r ON r.rental_id = t.rental_id
  JOIN inventory i ON i.inventory_id = r.inventory_id
  JOIN film f ON f.film_id = i.film_id
  WHERE t.total_amount = (
    SELECT MAX(total_amount)
    FROM (
      SELECT SUM(amount) AS total_amount
      FROM payment
      GROUP BY rental_id
    )
  )
  ORDER BY f.title;
  ")
}
c(SQL_c())
```

```{r}
bench_c <- microbenchmark(
  R = R_c(),
  SQL = SQL_c()
)
print(bench_c)
```

Both methods give the same result, and SQL is faster.

## **Problem 3 - Australian Records**

```{r}
data <- read.csv("Au-500.csv")
```

## (a).

```{r}
length(data$web[grepl("com$", data$web)])/nrow(data)
```

No website is .com's .

## (b).

```{r}
email <- data$email
domain <- tolower(trimws(sub("^[^@]*@", "", email)))
tab <- sort(table(domain), decreasing = TRUE)
kable(head(tab))
```

hotmail.com is the most common domain name amongst the email addresses.

## (c).

```{r}
company <- data$company_name
x1 <- gsub("[,[:space:]]", "", company)
nonalpha1 <- grepl("[^A-Za-z]", x1)
mean(nonalpha1)
```

```{r}
x2 <- gsub("[,[:space:]&]", "", company)
nonalpha2 <- grepl("[^A-Za-z]", x2)
mean(nonalpha2)
```

Excluding commas and whitespace, **9%** of company names contain a non-alphabetic character.

If we also exclude ampersands (“&”), the proportion drops to **0.8%**

## (d).

```{r}
d1  <- gsub("\\D", "", data$phone1)
ok1 <- nchar(d1) == 10
data$phone1 <- ifelse(ok1,
                      sub("^(\\d{4})(\\d{3})(\\d{3})$",
                          "\\1-\\2-\\3", d1),
                      data$phone1)

d2  <- gsub("\\D", "", data$phone2)
ok2 <- nchar(d2) == 10
data$phone2 <- ifelse(ok2,
                      sub("^(\\d{4})(\\d{3})(\\d{3})$",
                          "\\1-\\2-\\3", d2),
                      data$phone2)

cat("phone1 (first 10):\n"); print(head(data$phone1, 10))
cat("\nphone2 (first 10):\n"); print(head(data$phone2, 10))
```

## (e).

```{r}
addr <- data$address
m <- regexpr("\\d+\\s*$", addr)  
apt <- rep(NA_integer_, length(addr))
apt[m > 0] <- as.integer(regmatches(addr, m))
apt_clean <- apt[!is.na(apt) & apt > 0]
hist(log(apt_clean),
     main = "Histogram of log(apartment numbers)",
     xlab  = "log(apartment number)")
```

## (f).

```{r}
table(substr(as.character(apt_clean), 1, 1))
```

By examining the first digit of each apartment number, its distribution aligns more closely with a uniform distribution than a decreasing one. Consequently, this data clearly violates Benford's Law, appearing more like artificially generated numbers than naturally occurring real-world numerical data.
